Evolutionary rates of SARS-CoV-2

### 2022-07-31 manuscript/variant_rates.tex
allow root swap
fix bib

\begin{figure*}[tb]
    \includegraphics[width=\textwidth]{figures/20I_rtt.png}
    \caption{{\bf Divergence increases linearly with time (20I (Alpha)).} Each panel shows the number of within-clade mutations (total (A), amino acid changing (B), synonymous (C)) as a function of time.
    The blue line in panel A indicates the divergence cut-off, panels B\&C only show sequences that pass the divergence filter. Each panel also shows mean $\pm$ standard deviation and a weighted linear fit.
    \label{fig:rate_alpha}}
\end{figure*}

@@

\begin{figure}
    \includegraphics[width=0.5\textwidth]{figures/rate_progression.pdf}
    \caption[]{{\bf Divergence and evolutionary rates of different Nextstrain clades over time.} Synonymous rates estimates are stable in time and fluctuate around the rate estimates for between clades. Non-synonymous rate estimates are highest for clades 19A - 20C.
    \label{fig:rate_progression} }
\end{figure}

@@

Adaptive evolution, however, is much harder to account for properly and can lead to more dramatic distortions.
Since the number of sites that allow beneficial mutations is small, adaptive evolution tends to be very stochastic.
Unlike neutral evolution, the rate of adaptive evolution depends on the population size and the transmission process.
Epistatic interactions can further steepen the dependence on population size and stochasticity.

\subsection*{Diversification within variants}
The analysis of evolutionary rates above only considers the mean divergence within rates.

\bibliography{bib}
\end{document}

>>>>

\begin{figure*}[tb]
    \includegraphics[width=\textwidth]{figures/rtt/20I_rtt.png}
    \caption{{\bf Divergence increases linearly with time (20I (Alpha)).} Each panel shows the number of within-clade mutations (total (A), amino acid changing (B), synonymous (C)) as a function of time.
    The blue line in panel A indicates the divergence cut-off, panels B\&C only show sequences that pass the divergence filter. Each panel also shows mean $\pm$ standard deviation and a weighted linear fit. Analogous figures for all variants are included in the appendix.
    \label{fig:rate_alpha}}
\end{figure*}

@@

\begin{figure}
    \includegraphics[width=0.5\textwidth]{figures/rate_progression.pdf}
    \caption{{\bf Divergence and evolutionary rates of different Nextstrain clades over time.} Synonymous rates estimates are stable in time and fluctuate around the rate estimates for between clades. Non-synonymous rate estimates are highest for clades 19A - 20C.
    \label{fig:rate_progression} }
\end{figure}

@@

Adaptive evolution, however, is much harder to account for properly and can lead to more dramatic distortions.
Since the number of sites that allow beneficial mutations is small, adaptive evolution tends to be very stochastic.
Unlike neutral evolution, the rate of adaptive evolution depends on the population size and the transmission process.
Epistatic interactions can further steepen the dependence on population size and stochasticity.
Methods to infer emergence dates and locations of the early variants might therefore be biased or over-confident.

\subsection*{Diversification within variants}
The analysis of evolutionary rates above only considers the mean divergence within clades which, at least when variation is neutral, is independent of epidemiological parameters.
The structure of within clade diversity does contain information on the rate of spread.
A simple model for a growing variant is a super-critical branching process with growth rate $\alpha$ and an embedded mutation process.
Offspring of genomes with $i$ mutations contain $i+j$ mutations, where $j$ is a Poisson distributed number with mean $\mu t$ (mutation rate $\mu$ and generation time $t$).
The probability that offspring genomes are identical to their parents is $u = 1-e^{-\mu t}$, which for a generation time of $t=5days$ and a rate of $\mu = 15/year$ evaluates to $u=0.2$.
Since these branching process models are linear, them mean number of cases $n$ will increase exponentially with rate $\alpha$, while the number of genomes with $i$ mutations relative to the founder $m_i$ grows with rate $\alpha - u$ per generation.
\begin{equation}
    \begin{split}
        \frac{dn}{dt} &= \alpha n \\
        \frac{dm_0}{dt} &= (\alpha - u) m_0 \\
        \frac{dm_1}{dt} &= u m_0 (\alpha - u) m_1 \\
        \frac{dm_2}{dt} &= u m_1 (\alpha - u) m_2 \\
        \cdots
    \end{split}
\end{equation}
with solution $m_i = e^{(\alpha - u)t} \frac{(ut)^i}{i!}$.
These Poisson weights of mutation numbers are again only valid if mutations accumulate along many independent lineages.
In particular, this assumption is violated if some lineages spread systematically faster than others either because of epidemiological factors or the mutations they carry.

\bibliography{bib}
\appendix
\include{appendix_rtt}
\include{counts_appendix}
\end{document}
